A tool for designing digital filters for the Hankel and Fourier transforms in potential, diffusive, and wavefield modeling

### 2018-01-16 ManuscriptGeophysics2018.tex
Start acknowledgment
Clean preamble
Adjust empyscript version
Include feeback KK

% \documentclass[manuscript]{geophysics}
\documentclass[paper,twocolumn,twoside]{geophysics}
% \documentclass[paper]{geophysics}
% \documentclass[paper, revised]{geophysics}
% \documentclass[manuscript,revised]{geophysics}
%%fakesection ===    PACKAGES & DEFINITIONS    ===
% \usepackage{showframe}

%% ~ ADDITIONAL PACKAGES TO GEOPHYSICS.CLS
\DeclareGraphicsExtensions{.pdf,.png,.jpg}
\usepackage{color}
\usepackage{tabularx}
\usepackage{colortbl}
\usepackage{booktabs}
\usepackage[USenglish]{babel}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{amssymb, amsmath, amsfonts}
\usepackage{upquote}
\usepackage[pdftex, final]{hyperref}
\hypersetup{allcolors=blue, allbordercolors={0 0 .5}, colorlinks=true}

\usepackage{listings}

% Figure directory
\renewcommand{\figdir}{./figures}
\usepackage{xspace}
\newcommand{\empymod}{\texttt{empymod}\xspace}
\usepackage[strings]{underscore}

\definecolor{MyGray}{gray}{0.85}
\newcolumntype{Y}{>{\columncolor{MyGray}\raggedleft\arraybackslash}r}
\newcolumntype{W}{>{\raggedleft\arraybackslash}r}
\newcolumntype{A}{>{\columncolor{MyGray}\raggedleft\arraybackslash}X}
\newcolumntype{B}{>{\raggedleft\arraybackslash}X}

\hyphenation{iso-tro-pic pe-ne-tra-ting}

\newcommand{\rmk}[1]{\textbf{\color{red}{#1}}}
\newcommand{\mr}[1]{\mathrm{#1}}
\frenchspacing

@@

\title{A digital filter designing tool for the Hankel and Fourier transforms in
potential-, diffusive-, and wavefield modeling}

@@

\address{
\footnotemark[1]Instituto Mexicano del Petróleo,
Eje Central Lázaro Cárdenas Norte 152,
Col. San Bartolo Atepehuacan C.P. 07730,
Ciudad de México, México,
E-mail: \href{mailto:dieter@werthmuller.org}{Dieter@Werthmuller.org};
\footnotemark[2]Lamont-Doherty Earth Observatory,
305C Oceanography,
61 Route 9W, PO Box 1000,
Palisades NY 10964-8000 US,
E-mail: \href{mailto:KKey@ldeo.columbia.edu}{KKey@ldeo.columbia.edu};
\footnotemark[3]TU Delft,
Building 23,
Stevinweg 1 / PO-box 5048,
2628 CN Delft,
E-mail: \href{mailto:E.C.Slob@tudelft.nl}{E.C.Slob@tudelft.nl}.}

@@

Deva Prasad Ghosh proposed in his Ph.D. thesis \citep{PhD.70.Ghosh} a linear
filter method for the Hankel transform which revolutionized the computational
modeling of electromagnetic responses in the field of geophysical exploration.
If you use a code that calculates electromagnetic responses in the
wavenumber-frequency domain and transforms them to the space-frequency domain
chances are very high that it uses the \emph{digital linear filter} method
(DLF). \cite{GP.71.Ghosh} states that the idea is based on suggestions made
four decades earlier by \cite{PHY.33.Slichter} and \cite{GEO.40.Pekeris}, in
that ``\emph{the kernel function is dependent only on the layer parameters, and
  an expression relating it to the field measurements can be obtained by
mathematical processes.}'' However, until the introduction of DLF by Ghosh
these suggestions found no widespread use, probably not last because of the
missing computer power to calculate the filter coefficients. He further states
that credit goes to his Ph.D. supervisor Otto \cite{BK.68.Koefoed,
GP.70.Koefoed}, who retook the task of direct interpretational methods with the
introduction of raised kernel functions. DLF is, as such, an improvement of
that approach, providing a faster and simpler method.

The introduction of linear filters to electromagnetic geophysics initiated a
wealth of investigation, publications, and development of computer programs
that extended and improved the method. A few dozen articles have been published
in the 70s and 80s by Ghosh, Koefoed, Das, Anderson, and many others, the vast
majority of it in the journal \emph{Geophysical Prospecting}. Another common
name of DLF is \emph{fast Hankel transform} (FHT), popular because of the
similarity of the name to the well known \emph{fast Fourier transform} (FFT).
The name was introduced, as far as we can see, by \cite{GP.79.Johansen} in
their article of the same name. However, the name FHT can be misleading as it
has \emph{Hankel} in the name, but the DLF can also be applied to the Fourier
transform (and other linear transforms). Publications fall broadly into one or
several of three categories: (1) new applications (extend the usage of DLF to
other measurement techniques); (2) filter improvements (new filters or
providing new or improved methods for the determination of filter
coefficients); and (3) computational tools. The following is a brief review of
the most relevant publications regarding DLF, without claiming completeness.

@@

Ghosh used the method originally for the computation of type curves for
Schlumberger and Wenner resistivity soundings. He published the main results of
his thesis the following year in two publications: in \cite{GP.71.Ghosh} he
derives a resistivity model from given Schlumberger or Wenner sounding curves;
and in \cite{GP.71.Ghoshb} he provides filters for the inverse operation,
deriving resistivity sounding curves from a given resistivity model.  The
method was next applied to electromagnetic soundings with horizontal and
perpendicular coils \citep{GP.72.Koefoed}, to vertical coplanar coil systems
\citep{GP.73.Verma}, to dipoles and other two electrode systems
\citep{GP.74.Das, GP.74.Dasb, GP.80.Das, GEO.94.Sorensen}, and to vertical
dikes, hence vertical instead of horizontal layers \citep{GEO.75.Niwas}. Whilst
the first filters were very specific to one type of curves and one type of
transformation, various publications used the method to get one type curve from
another type curve \citep{GP.77.Kumar, GP.78.Kumar} or generalized the method
to be applicable to a wider set of problems \citep{EXG.80.Davis, GXP.81.Das,
GEO.84.Das, GP.84.ONeill}. Eventually, it passed from pure layered modeling to
primary-secondary formulations in 3D problems, where DLF is used to compute the
spatial Fourier-Hankel transforms in a horizontally layered background medium
and to compute transient responses from frequency domain computations
\citep{GJI.81.Das, GJI.82.Das, GEO.84.Anderson, GEO.86.Newman,
MGS.17.Kruglyakov}. Various authors delved into the theory of the method,
analyzing the oscillating behaviour of the filters and trying to estimate the
error of DLF \citep{GP.72.Koefoedb, GP.76.Koefoedb, GP.79.Johansen,
GP.90.Christensen}.

@@

\cite{PhD.70.Ghosh} derived the filter coefficients in the spectral domain by
dividing the output spectrum by the input spectrum followed by an inverse
Fourier transform. Improvements to the determination of filter coefficients
were provided by \cite{EXG.75.ONeill, GEO.77.Nyman, GEO.82.Das}, or
specifically for the Fourier transform by \cite{GP.86.Nissen}. A direct
integration method was used by \cite{GP.76.Bichara, GP.78.Bernabini}.
\cite{GP.79.Koefoed} proposed a Wiener-Hopf least-squares method, which was
further improved by many authors \citep{GP.82.Guptasarma, GEO.1982.Murakami,
GP.97.Guptasarma}. \cite{GP.07.Kong} proposed a direct matrix inversion method
to solve the convolution equation, which requires only the input and output
sample values. To evaluate the filters he defines the criteria of a good filter
as one that recovers small or weak diffusive EM fields. This method was also
used by \cite{GEO.09.Key, GEO.12.Key}. Most authors publish filters for the
Hankel transform with $J_0$ and $J_1$ Bessel functions (or $J_{-1/2}$,
$J_{1/2}$ if applied to the Fourier sine/cosine transform), as all higher
Bessel functions can be rewritten to only use these two. \cite{GP.94.Mohsen} is
one of the rare cases which provides $J_2$ filter weights.

@@

The best known codes are likely the freely available ones by Anderson.
\cite{USGS.73.Anderson} extends the method to transient responses, applying DLF
not only to the Hankel transform, but also to the Fourier transform. A
transient signal can therefore be obtained be applying twice a digital filter
to the wavenumber-frequency domain calculation. In \cite{USGS.75.Anderson,
GEO.79.Anderson} he presents improved filters for both Fourier and Hankel
transforms, introducing measures to significantly speed-up the calculation,
such as the lagged convolution or using the same abscissae for $J_0$ and $J_1$.
\cite{TMS.82.Anderson} and the included 801\,pt filter became sort of the
industry standard, to which subsequent filters were compared. In
\cite{GEO.89.Anderson} he presents a hybrid solution, permitting to use either
DLF or quadrature and as such permits to compare the two. Other examples
include the codes by \cite{GP.75.Johansen}, an interactive system for
interpretation of resistivity soundings, and a tool to calculate filter
coefficients by \cite{GP.90.Christensen}. The latter is available upon request
and was used, for instance, in all the open-source modeling and inversion
routines of CSIRO in the Amira Project 223 \citep{ASEG.07.Raiche}. \newline

All mentioned publications have in common that they were derived for direct
current methods (DC) or low frequency methods, such as time-domain shallow EM
methods (TEM), or controlled-source electromagnetics (CSEM), but not for high
frequency methods such as ground-pe\-ne\-tra\-ting radar (GPR). Generally it
was even thought that the filter method works only in the diffusive limit
\citep[e.g.,][]{GEO.15.Hunziker}. Also, there is no open-source filter
designing tool readily available, they are available as a described method
in articles or as code upon request by the author, if at all. The presented
algorithm \texttt{fdesign} tries to fill this gap by providing the tools to
design general or purpose-built filters using the direct matrix inversion
method. The current article falls therefore into all three mentioned categories
by applying DLF also to wave phenomena and providing new filters with the
accompanying code. After a brief overview of the methodology, the theory, and
the code we show examples of its usage, presenting some new filters for both,
CSEM and GPR data. The algorithm and many more examples of its usage can be
found on
\href{https://github.com/empymod/article-fdesign}{github.com/empymod/article-fdesign}.
The examples can be used as templates to design new filters.

@@

The algorithm \texttt{fdesign} is a filter designing tool using the direct
matrix inversion method as described by \cite{GP.07.Kong} and based on scripts
by \cite{GEO.12.Key}. The tool is an add-on to the electromagnetic modeler
\texttt{empymod} \citep{GEO.17.Werthmuller}, written in Python, and hosted on
GitHub, which should foster interaction and enable anyone to toy around,
improve, and extend it. It can be used to derive digital linear filters for the
Hankel and the Fourier transforms for potential-, diffusive-, and wavefields
(hence from DC to GPR). Theoretically, it can be used to derive linear filters
for any linear transform, as long as you feed it with a theoretical, or
accurately computed, transform pair of the transform. It permits to derive
filters with the help of theoretical transform pairs or, alternatively, with
the electromagnetic modeler \texttt{empymod}, using a quadrature method to
derive accurate curves used for the inversion.

The main points of the method and some differences between it and the methods
it is based upon are:
\begin{itemize}
  \item The algorithm computes many different filters for various spacing and
    shift values (brute-force), with an optional minimization routine for
    the best outcome of the brute-force step.
  \item \cite{GP.07.Kong, GEO.12.Key} optimize a filter for the $J_0$ filter,
    and then use the obtained shift and spacing value to calculate the $J_1$
    values. The presented algorithm can optimize either $J_0$, $J_1$, or both
    in the optimization process. Five transform-pairs for each $J_0$ and $J_1$,
    and three transform pairs for each sine and cosine of often used functions
    are included in the routine, but any other transform pair can be provided
    as input.
  \item More complex or more specific models can be provided instead of
    theoretical transform pairs by using the modeler \texttt{empymod}.
  \item \cite{GP.07.Kong} defines a \emph{good} filter as one that recovers
    \emph{weak} diffusive EM fields. In the presented algorithm you can define
    a relative error level which defines up to what error an obtained result is
    good or not. The presented algorithm can also minimize the amplitude, but
    it has additionally another mode, where you maximise the abscissae $r$ of
    the right-hand-side of the transform pair instead of minimising the
    amplitude. The two modes yield the same result in the general, simple case
    of fast decaying transform pairs. However, if you use complex models for
    the design, and specifically high frequencies, then maximising $r$ will
    yield much better and more consistent results.
  \item The algorithm allows to run under-, equal-, and over-determined
    systems.
  \item The real or the imaginary part can be used for the inversion of complex
    transform pairs.
\end{itemize}

The quality of a filter depends heavily on the model chosen for comparison. An
obtained filter might be very good for one model, but not that good for another
one, which is no different for this designing tool. There is no way to estimate
the error of a result obtained with a certain filter if you apply it to any
other model, unless you calculate this other model with another method for
comparison. All results presented here should therefore be taken with a certain
care.

@@

Most of the articles mentioned in the review have detailed derivations of the
digital filter method. In this article we focus on the algorithm, and summarize
the theory only very briefly by following \cite{GEO.12.Key}. In
electromagnetics we often have to evaluate integrals of the form
%
\begin{equation}
  F(r) = \int^\infty_0 f(l)K(l r)\mr{d}l \ ,
  \label{eq:HankelInt}
\end{equation}
%
where $l$ and $r$ denote left-hand-side and right-hand-side evaluation values,
and $K$ is the kernel function. In the specific case of the Hankel transform
$l$ corresponds to wavenumber, $r$ to offset, and $K$ to Bessel functions; in
the case of the Fourier transform $l$ corresponds to frequency, $r$ to time,
and $K$ to sine or cosine functions. In both cases it is an infinite integral
which numerical integration is very time-consuming because of the slow decay
of the kernel function and its oscillatory behaviour.

By substituting $r = e^x$ and $l = e^{-y}$ we get
\begin{equation}
  e^x F(e^x) = \int^\infty_{-\infty} f(e^{-y})K(e^{x-y})e^{x-y}\mr{d}y\ .
  \label{eq:filtint}
\end{equation}
This can be re-written as a convolution integral and be approximated for an
$N$-point filter by
\begin{equation}
  F(r) \approx \sum^N_{n=1} \frac{f(b_n/r) h_n}{r}\ ,
  \label{eq:filtapprox}
\end{equation}
where $h$ is the digital linear filter, and the logarithmically spaced filter
abscissae is a function of the spacing $\Delta$ and the shift $\delta$,
\begin{equation}
  b_n = \exp\left\{\Delta(-N/2+n) + \delta\right\} \ .
  \label{eq:base}
\end{equation}
From equation~\ref{eq:filtapprox} it can be seen that the filter method
requires $N$ evaluations at each $r$. To calculate for instance the frequency
domain result for 100 offsets with a 201\,pt filter requires 20'100 evaluations
in the wavenumber domain. This is why the DLF often uses interpolation to
minimise the required evaluations, either in the right-hand side (often called
\emph{lagged convolution DLF}) or in the left-hand side (\emph{splined DLF}).

\subsection{Pseudo-code}
The main input variables are the filter length ($N$), the spacing ($\Delta$)
and shift ($\delta$) ranges over which to loop, and the transform pairs for
the inversion ($f_\mr{I}$) and the check of quality ($f_\mr{C}$). If
$f_\mr{C}$ is not provided, then $f_\mr{I}$ is used for both. There are
additional, optional input parameters, for instance to adjust how the rhs
abscissae $r = f(b, N)$ are calculated, where $b$ is the filter base.

The basic steps are as follows:
\begin{enumerate}
  \item Evaluate rhs of check-function $f_\mr{C}$:\newline
        $d_\mr{R} = f_\mr{C}.\mr{rhs}(r)$
  \item Loop over each value $\Delta_i$, $\delta_j$ (brute force):
    \begin{enumerate}
      \item Calculate filter base:\newline
        $b_n = \exp\left\{\Delta_i(-N/2+n) + \delta_j\right\}$
      \item Get required rhs ($r$) and lhs ($l$) evaluation points:\newline
        $r = f(b, N)$\newline
        $l = b/r$
      \item Invert for filter coefficients:\newline
        $J_{ij} = \mr{solve}(f_\mr{I}.\mr{lhs}(l), f_\mr{I}.\mr{rhs}(r))$
    \item Calculate numerically rhs of check-function $f_\mr{C}$ with current
      filter $J_{ij}$:\newline
      $d_\mr{F} = f_\mr{C}.\mr{lhs}(l)\cdot J_{ij}\ /\ r$
      \item Store minimum recovered amplitude or maximum $r$ where relative
        error is less than the provided error:\newline
        $\chi_{ij} = g\left[
        \mr{argmin}\left(|(d_\mr{F}-d_\mr{R})/d_\mr{R}| >
        \mr{error}\right) -1\right]\ $, where $g$ is either $d_\mr{R}$
        or $1/r$.
    \end{enumerate}
  \item Return filter coefficients which yield minimum amplitude or maximum $r$
    (a local minimization can be run to polish the brute-force result):\newline
    return $J\left[\mr{argmin}(\chi_{ij})\right] $\newline
\end{enumerate}

>>>>

% \documentclass[manuscript]{geophysics}
\documentclass[paper,twocolumn,twoside]{geophysics}
% \documentclass[paper]{geophysics}
% \documentclass[paper, revised]{geophysics}
% \documentclass[manuscript,revised]{geophysics}

%%fakesection ===    PACKAGES & DEFINITIONS    ===
% \usepackage{showframe}

%% ~ ADDITIONAL PACKAGES TO GEOPHYSICS.CLS
\usepackage[USenglish]{babel}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{amssymb, amsmath, amsfonts}
\usepackage{upquote}
\usepackage[strings]{underscore}
\usepackage[pdftex, final]{hyperref}
\hypersetup{allcolors=blue, allbordercolors={0 0 .5}, colorlinks=true}

% Figures
\DeclareGraphicsExtensions{.pdf,.png,.jpg}
\renewcommand{\figdir}{./figures}

% Own commands
\usepackage{xspace}
\newcommand{\empymod}{\texttt{empymod}\xspace}
\newcommand{\mr}[1]{\mathrm{#1}}

@@

\title{A tool for designing digital filters for the Hankel and Fourier transforms in
potential, diffusive, and wavefield modeling}

@@

\address{
\footnotemark[1]Instituto Mexicano del Petróleo,
Eje Central Lázaro Cárdenas Norte 152,
Col. San Bartolo Atepehuacan C.P. 07730,
Ciudad de México, México,
E-mail: \href{mailto:dieter@werthmuller.org}{Dieter@Werthmuller.org};
\footnotemark[2]Lamont-Doherty Earth Observatory, Columbia University,
305C Oceanography,
61 Route 9W, PO Box 1000,
Palisades NY 10964-8000 US,
E-mail: \href{mailto:KKey@ldeo.columbia.edu}{KKey@ldeo.columbia.edu};
\footnotemark[3]TU Delft,
Building 23,
Stevinweg 1 / PO-box 5048,
2628 CN Delft,
E-mail: \href{mailto:E.C.Slob@tudelft.nl}{E.C.Slob@tudelft.nl}.}

@@

In his Ph.D. thesis, \cite{PhD.70.Ghosh} proposed a linear filter method for
the numerical evaluation of Hankel transforms that subsequently revolutionized
the computation of electromagnetic (EM) responses in the field of geophysical
exploration. If you use a code that calculates EM responses in the
wavenumber-frequency domain and transforms them to the space-frequency domain,
chances are high that it uses the \emph{digital linear filter} (DLF) method.
Most practical 1D EM modeling codes rely on the DLF method for rapid
computations; these codes are not only used for standalone simulations of EM
fields in layered 1D media, but they are also commonly embedded within 3D
modeling codes for computing the primary fields in scattered-field formulations
or for the background fields required by integral equation methods. Thus, the
DLF method is an important component of many commonly used modeling codes for
EM geophysical data.

\cite{GP.71.Ghosh} states that the DLF idea is based on suggestions made four
decades earlier by \cite{PHY.33.Slichter} and \cite{GEO.40.Pekeris}, in that
``\emph{the kernel function is dependent only on the layer parameters, and an
expression relating it to the field measurements can be obtained by
mathematical processes.}'' However, until the introduction of DLF, these
suggestions found no widespread use, likely because of the missing computer
power to calculate the filter coefficients. He further states that credit goes
to his Ph.D. supervisor Otto \cite{BK.68.Koefoed, GP.70.Koefoed}, who retook
the task of direct interpretational methods with the introduction of raised
kernel functions. DLF is, as such, an improvement of that approach, providing a
faster and simpler method.

The DLF technique is sometimes referred to as the \emph{fast Hankel transform}
(FHT), popular because of the similarity of the name to the well known
\emph{fast Fourier transform} (FFT), although the algorithms for these
techniques are completely different. The FHT name was likely inspired by the
title of a paper by \cite{GP.79.Johansen}. However, the name FHT can be
misleading as it has \emph{Hankel} in the name, whereas the DLF approach can
more generally be applied to other linear transforms, for example Fourier sine
and cosine transforms.

The introduction of linear filters to EM geophysics, in parallel with rising
computational power, initiated a wealth of investigations, leading to the
development of computer programs that extended and improved the DLF method, and
to numerous publications. These publications fall broadly into one or several
of three categories: (1) new applications that extend the usage of DLF to other
EM measurement techniques; (2) filter improvements that provide either new
filters or improved methods for the determination of filter coefficients; and
(3) computational tools that compute EM responses using DLF techniques. Here,
we briefly review the most relevant publications, without claiming
completeness.

@@

Ghosh used the method originally for the computation of type curves for
Schlumberger and Wenner resistivity soundings: \cite{GP.71.Ghosh} derives a
resistivity model from given Schlumberger or Wenner sounding curves; and
\citet{GP.71.Ghoshb} provides filters for the inverse operation, deriving
resistivity sounding curves from a given resistivity model. The method was next
applied to electromagnetic soundings with horizontal and perpendicular coils
\citep{GP.72.Koefoed}, to vertical coplanar coil systems \citep{GP.73.Verma},
to dipoles and other two electrode systems \citep{GP.74.Das, GP.74.Dasb,
GP.80.Das, GEO.94.Sorensen}, and to vertical dikes, hence vertical instead of
horizontal layers \citep{GEO.75.Niwas}. The first filters were specific to a
particular resistivity sounding type and its transform; later publications used
the method to get one type curve from another type curve \citep{GP.77.Kumar,
GP.78.Kumar}, or generalized the method to be applicable to a wider set of
problems \citep{EXG.80.Davis, GXP.81.Das, GEO.84.Das, GP.84.ONeill}.
Eventually, it passed from pure layered modeling to primary-secondary field
formulations for 3D problems, where DLF is used to compute the spatial
Fourier-Hankel transforms in a horizontally layered background medium and to
compute transient responses from frequency domain computations
\citep{GJI.81.Das, GJI.82.Das, GEO.84.Anderson, GEO.86.Newman,
MGS.17.Kruglyakov}. Other publications delved into the theory of the method,
analyzing the oscillating behaviour of the filters and trying to estimate the
error of DLF \citep{GP.72.Koefoedb, GP.76.Koefoedb, GP.79.Johansen,
GP.90.Christensen}.

@@

\cite{PhD.70.Ghosh} derived the filter coefficients in the spectral domain by
dividing the output spectrum by the input spectrum followed by an inverse
Fourier transform. Improvements to the determination of filter coefficients
were provided by \cite{EXG.75.ONeill, GEO.77.Nyman, GEO.82.Das}, or
specifically for the Fourier transform by \cite{GP.86.Nissen}. A direct
integration method was used by \cite{GP.76.Bichara} and \cite{GP.78.Bernabini}.
\cite{GP.79.Koefoed} proposed a Wiener-Hopf least-squares method, which was
further improved by many authors \citep{GP.82.Guptasarma, GEO.1982.Murakami,
GP.97.Guptasarma}. \cite{GP.07.Kong} proposed a direct matrix inversion method
to solve the convolution equation, which requires only the input and output
sample values. To evaluate the filter's effectiveness, he defines a good filter
as one that recovers small or weak diffusive EM fields. This method was also
used by \cite{GEO.09.Key, GEO.12.Key} to create a suite of filters accurate for
marine EM data. Most works have published filters for the Hankel transform with
$J_0$ and $J_1$ Bessel functions (or $J_{-1/2}$, $J_{1/2}$ if applied to the
Fourier sine/cosine transform), as all higher Bessel functions can be
rewritten, via recurrence relations, using only use these two.
\cite{GP.94.Mohsen} is one of the rare cases which provides $J_2$ filter
weights.

@@

The most well-known codes are likely Anderson's freely available ones.
\cite{USGS.73.Anderson} extends the method to transient responses, applying DLF
not only to the Hankel transform, but also to the Fourier transform. A
transient signal can therefore be obtained be applying twice a digital filter
to the wavenumber-frequency domain calculation. \cite{USGS.75.Anderson,
GEO.79.Anderson} presents improved filters for both Fourier and Hankel
transforms, introducing measures to significantly speed-up the calculation,
such as the lagged convolution or using the same abscissae for $J_0$ and $J_1$.
\cite{TMS.82.Anderson} and the included 801\,pt filter became the de facto
industry standard, to which subsequent filters were compared.
\cite{GEO.89.Anderson} presents a hybrid solution that uses either the DLF or a
much slower but highly accurate adaptive-quadrature approach, which can also be
used to measure the accuracy of the DLF. \cite{GEO.12.Key} presented codes for
comparing the DLF approach with a speed optimized quadrature method call
quadrature-with-extrapolation (QWE); despite its optimizations, QWE is still
not as fast as DLF, but it remains valuable as an independent tool for testing
the accuracy of particular filters for the DLF method. Other examples include
the codes by \cite{GP.75.Johansen}, an interactive system for interpretation of
resistivity soundings, and a tool to calculate filter coefficients by
\cite{GP.90.Christensen}. The latter is available upon request and was used,
for instance, in all the open-source modeling and inversion routines of CSIRO
in the Amira Project 223 \citep{ASEG.07.Raiche}.

All mentioned publications have in common that they were derived for direct
current methods (DC) or low frequency methods, such as time-domain shallow EM
methods (TEM), or controlled-source electromagnetics (CSEM), but not for high
frequency methods such as ground-pe\-ne\-tra\-ting radar (GPR). Generally it
was even thought that the filter method works only in the diffusive limit where
the quasistatic approximation is valid \citep[e.g.,][]{GEO.15.Hunziker}.
Nevertheless, \cite{GEO.17.Werthmuller} tested DLF for modeling 250 MHz
center-frequency GPR data using a 401 point filter that was designed for
diffusive EM, and found that it was orders of magnitude faster than quadrature
approaches. However, good quality results were only obtained for the first
arrivals at short time-offsets and the later arrivals had poor quality. Yet,
this promising initial test suggested it might be possible to specifically
design a special filter for GPR frequencies where the EM fields propagate as
waves rather than by diffusion.

While there are several versions of digital filters that are freely available,
there are no open-source filter designing tools that readily available; they
are either only described methods in articles or the codes are only available
upon request by the author. Here, we address this gap by presenting the code
\texttt{fdesign}, which is an open-source tool for designing general or
purpose-built filters using the direct matrix inversion method. The code is
completely open-source and can be run using freely available programming
environments, thus there are few barriers to obtaining and running the code.

In the following sections we give a brief overview of the methodology, the
theory, and the code, and then show examples of its usage, as well as
presenting new filters for both CSEM and GPR data. The current article falls
therefore into all three mentioned categories by the new application of DLF to
wave phenomena and by providing new filters with the accompanying code. The
algorithm and many more examples of its usage can be found on
\href{https://github.com/empymod/article-fdesign}{github.com/empymod/article-fdesign}.
The examples can be used as templates to design new filters.

@@

The algorithm \texttt{fdesign} is a filter designing tool using the direct
matrix inversion method as described by \cite{GP.07.Kong} and based on scripts
developed by \cite{GEO.12.Key}. The tool is an add-on to the EM modeler
\texttt{empymod} \citep{GEO.17.Werthmuller}, written in Python, and hosted on
GitHub, which should foster interaction and enable anyone to toy around,
improve, and extend it. It can be used to derive digital linear filters for the
Hankel and the Fourier transforms for potential, diffusive, and wavefield
modeling (hence from DC to GPR).

Theoretically, it can be used to derive linear filters for any linear
transform, as long as it is supplied with the input and outputs of a transform
pair. Previous works have relied on using theoretical transforms pairs commonly
found in tables of integral transforms. Here we follow a similar approach, but
we also propose a new method where filters could be derived from accurately
computed numerical transform pairs, such as those computed using quadrature
approaches. For example, the EM modeler \texttt{empymod} could be used to
compute transform pairs for a particular class of model and data parameters
(for example, a specific EM method and model of interest) using a slow but
highly accurate quadrature method. A good filter derived from the resulting
transform pairs could then be used for faster modeling of similar model and
data parameters.

The main features of \texttt{fdesign} and some differences between it and the
methods it is based upon are:
\begin{itemize}
  \item The algorithm computes many different filters for various spacing and
    shift values (brute-force), with an optional minimization routine for
    the best outcome of the brute-force step.
  \item \cite{GP.07.Kong} and \cite{GEO.12.Key} optimize a filter for the $J_0$
    filter, and then use the obtained shift and spacing value to calculate the
    $J_1$ values. The presented algorithm can optimize either $J_0$, $J_1$, or
    both in the optimization process. Five transform-pairs for each $J_0$ and
    $J_1$, and three transform pairs for each sine and cosine of often used
    functions are included in the routine, but any other transform pair can be
    provided as input.
  \item More complex or more specific numerical transforms  pairs  can be
    provided instead of theoretical transform pairs by using the modeler
    \texttt{empymod}.
  \item \cite{GP.07.Kong} defines a \emph{good} filter as one that recovers
    \emph{weak} diffusive EM fields. In  \texttt{fdesign}, you can define a
    relative error level that defines up to what error an obtained result is
    good or not. The optimization seeks to find smallest amplitude field that
    is accurately modeled (i.e., minimizing the amplitude), but it has
    additionally another mode, where it will maximize the abscissae $r$ of the
    right-hand-side of the transform pair instead of minimizing the amplitude.
    The two modes yield the same result in the general, simple case of fast
    decaying transform pairs. However, for some complex models and specifically
    for very high frequencies, maximizing $r$ will yield much better and more
    consistent results.
  \item The algorithm can solve under-, equal-, and over-determined linear
    systems.
  \item The real or the imaginary part can be used for the inversion of complex
    transform pairs.
\end{itemize}

The quality of a filter depends heavily on the model chosen for comparison. An
obtained filter might be very good for one model or data acquisition setup, but
not that good for another one, which is no different for this designing tool.
There is no way to estimate the error of a result obtained with a certain
filter if you apply it to any other model, unless you calculate this other
model with another method for comparison. All results presented here should
therefore be taken with a certain care. Further, when applying a new filter to
a significantly different model and/or data space than what it has previously
been tested on, we recommend first verifying its accuracy in that space by a
comparison with results computed with a quadrature method such as QWE, which is
also available in \texttt{empymod}.

@@

Most of the articles mentioned in the review have detailed derivations of the
digital filter method. In this article we focus on the algorithm, and summarize
the theory only very briefly by following \cite{GEO.12.Key}. In
electromagnetics we often have to evaluate integrals of the form
%
\begin{equation}
  F(r) = \int^\infty_0 f(l)K(l r)\mr{d}l \ ,
  \label{eq:HankelInt}
\end{equation}
%
where $l$ and $r$ denote evaluation values, and $K$ is the kernel function. In
the specific case of the Hankel transform $l$ corresponds to wavenumber, $r$ to
offset, and $K$ to Bessel functions; in the case of the Fourier transform $l$
corresponds to frequency, $r$ to time, and $K$ to sine or cosine functions. In
both cases it is an infinite integral which numerical integration is very
time-consuming because of the slow decay of the kernel function and its
oscillatory behaviour.

By substituting $r = e^x$ and $l = e^{-y}$ we get
%
\begin{equation}
  e^x F(e^x) = \int^\infty_{-\infty} f(e^{-y})K(e^{x-y})e^{x-y}\mr{d}y\ .
  \label{eq:filtint}
\end{equation}
%
This can be re-written as a convolution integral and be approximated for an
$N$-point filter by
%
\begin{equation}
  F(r) \approx \sum^N_{n=1} \frac{f(b_n/r) h_n}{r}\ ,
  \label{eq:filtapprox}
\end{equation}
%
where $h$ is the digital linear filter, and the logarithmically spaced filter
abscissae is a function of the spacing $\Delta$ and the shift $\delta$,
%
\begin{equation}
  b_n = \exp\left\{\Delta(-N/2+n) + \delta\right\} \ .
  \label{eq:base}
\end{equation}
%
From equation~\ref{eq:filtapprox} it can be seen that the filter method
requires $N$ evaluations at each $r$. For example, to calculate the frequency
domain result for 100 offsets with a 201\,pt filter requires 20,100 evaluations
in the wavenumber domain. This is why the DLF often uses interpolation to
minimize the required evaluations, either for $F(r)$ in what is referred to as
\emph{lagged convolution DLF}, or for $f(l)$, which we here call \emph{splined
DLF}.

In the direct matrix inversion method for solving for the digital filter
coefficients, equation \ref{eq:filtapprox} is cast as a linear system as
follows. $\Delta$ and $\delta$ are preassigned scalar values and $r$ is a range
of $M$ preassigned values $r_m$. The filter base coefficients $b_n$ are
computed using equation \ref{eq:base} and an array of values for $l$ are
computed using $l_{mn} = b_n/r_m$. The linear system's left hand side (LHS)
matrix $\bf{A}$ has dimensions $M \times N$ with coefficients $A_{mn} =
f(l_{mn})/r_n$ and the right hand side (RHS) vector $\bf{v}$ has elements $v_n
= F(r_n)$. The resulting linear system
%
\begin{equation}
  \bf{A}\bf{h} = \bf{v}
  \label{eq:,linearSystem}
\end{equation}
%
can be solved by direct matrix inversion to obtain the vector of filter
coefficients $\bf{h}$. For a given filter length $N$, there are many subjective
choices that go into the practical implementation of this method, including the
choice of the transform pairs $F(r)$ and $f(l)$, as well as the value for $M$
and the particular range and spacing of the values $r_m$. Once values are
chosen for these variables, an optimal filter can be found by a grid search
over $\Delta$ and $\delta$ for values that produce a high quality filter. The
choice of metric for what constitutes a high quality filter is also subjective
and is further discussed below.

\subsection{Pseudo-code for \texttt{fdesign} }

The main input variables are the filter length ($N$), the spacing ($\Delta$)
and shift ($\delta$) ranges over which to loop, and the transform pairs for the
inversion ($f_\mr{I}$) and the check of quality ($f_\mr{C}$). If $f_\mr{C}$ is
not provided, then $f_\mr{I}$ is used for both. There are additional, optional
input parameters, for instance to adjust how the RHS abscissae for the
inversion $r_\mr{I} = p(b, N)$ are calculated, where $b$ is the filter base.
Additionally, ranges $r$ for evaluating the check function need to be
specified.
